On the Minimum Core Mass for Giant Planet Formation

### 2013-10-17 tex/draft/draft4.tex
section 4 edits      Big problem was that luminosity explanation was still wrong/incomplete. Also claims that ignoring L_negl is conservative don't seem rigorous and were removed as unneeded anyway. Fig. 5 caption was both long and incomplete (no mention of RCB/negative values).

Deep in a non-self-gravitating atmosphere, \Eq{eq:deep} shows that $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$.  This behavior is seen in the low mass solutions in \Fig{fig:profiles}.  As explained by \Eq{eq:TPsg}, the higher mass solutions show a flatter profile in $T$ and also in $P \propto T^{1/\delad}$.  In agreement with \Eq{eq:cb2}, the radiative zone remains nearly isothermal -- even in the more massive solutions -- and thus shows a nearly exponential increase in pressure with inverse depth.

The atmosphere starts out as purely convective and isentropic, as seen in \Fig{fig:profiles} for the 5.01 $M_{\oplus}$ solution which has no radiative zone. Further gas accumulation brings in energy that has to be radiated away, forming a radiative region that quickly deepens. The fast pressure deepening of the radiative region is evident in \Fig{fig:profiles} --- even a low mass atmosphere (i.e. the 5.10 $M_{\oplus}$ solution) has an extended radiative zone.

The higher mass atmosphere solutions have lower entropy, i.e.\ cooling allows the atmosphere to accrete more gas.  This is evident in \Fig{fig:profiles} from the higher internal pressures of the higher mass solutions.  The somewhat higher temperatures in the higher mass solutions do not cause the entropy to increase.

%AY: to say something about starting at disk entropy (not really needed) we should show it better in the figure.  Also unclear if you want to say much more about evolution before figure 2 shows the evolution.
%In the initial stages of its evolution, the atmosphere is purely convective and isentropic, with an entropy equal to that of the disk. In the context of our numerical model, this is the minimum atmosphere mass for which a hydrostatic solution exists. Further gas accumulation brings in energy that has to be radiated away, and an initially thin outer radiative region forms, which becomes thicker as the atmosphere becomes more massive.

@@

The cooling model of \S\ref{sec:twolayer} is used to connect solutions of different atmospheric masses into an evolutionary sequence.  \Fig{fig:Ltplot} shows the luminosity evolution and the elapsed time as a function of atmospheric mass for a $5 M_{\oplus}$ core at $60$ AU, i.e.\ the same parameters as in \Fig{fig:profiles}.  During the early stages of atmospheric growth, the luminosity drops sharply.  This behavior is seen in both the full numerical solutions (with solid lines) and the analytic model (with dashed lines).  As seen in \Eq{eq:Lcb}, the increased pressure depth of the RCB increases the optical depth ($\propto \kappa P)$ and decreases the radiative luminosity.

%AY: I don't want to say 5.12 is the lowest mass b/c it's not the adiabat, and I don't see what's gained by saying it's very near the lowest mass.

At later stages of evolution, the numerical model gives a flat luminosity with increasing mass and also time (not shown).  By contrast the analytic model gives a luminosity that continues to drop with increasing mass.  The difference is due to the neglect of self-gravity in the analytic model.  From \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$.  Thus accounting for the higher enclosed mass gives a somewhat higher luminosity in the self-gravitating model.  However, the main effect is the higher temperature in the self-gravitating models.  The flatter temperature profiles with radius, as described above, explain why the self-gravitating models are hotter.

%The flattening of the luminosity curve is a generic feature of numerical core accretion models

%We find only a brief agreement between the analytic and numerical results. In the early stages, the radiative zone is too shallow and the assumption that $P_{\rm RCB} \gg P_{\rm d}$ used in the analytic model breaks down.

Before discussing time evolution (both the elapsed time from the bottom panel of \Fig{fig:Ltplot} and the characteristic growth time of the atmosphere), we briefly discuss the effect of decreasing the atmospheric dust abundance on the atmospheric evolution.  In \Fig{fig:LtvsMopacity}, the opacity normalization in \Eq{eq:opacitylaw} is reduced by factors of 10 and 100.  Lower opacities result in higher luminosities and faster evolution, a well-established result \citep{HubBod05} that we confirm for our model.  Atmospheric dust opacities are difficult to robustly predict. Ablation of infalling solids is a dust source, while settling through the radiative zone is a sink.  Grain growth both reduces dust opacities per unit mass and favors settling.  Our scenario of negligible ongoing particle accretion tends to favor low dust opacities.   To be conservative, however, our reference case considers full Solar abundances, but faster evolution is clearly possible. The effect of opacity reduction on the critical core mass is deferred to \S\ref{sec:critical}.

@@

\Fig{fig:growthtime} plots the evolution of the atmospheric growth timescale, $M_{\rm atm}/\dot{M}$,  around a $5 M_{\oplus}$ core at several locations in the disk.  This instantaneous growth time shows clearly that the atmosphere spends the bulk of its time growing though intermediate atmospheric masses, $\sim 1 -3$ $M_\oplus$ in this case.  Growth times are short both early -- when the radiative zone is relatively transparent -- and late -- when self-gravity accelerates growth.  The accelerated growth at higher masses is also evident in \Figs{fig:Ltplot}{fig:LtvsMopacity} (bottom panels) as the flattening of the $t$ vs.\ $M_{\rm atm}$ curve.

The fact that growth times peak before the crossover mass, when $M_{\rm{atm}} < M_{\rm c}$, is very helpful for understanding atmospheric growth.  The estimated timescale to runaway growth can be well approximated by only considering the time it takes to grow to the crossover mass, or even to a somewhat lower mass.  This insensitivity to the upper mass threshold is characteristic of accelerating growth, and is fortuitious.  We next show how our model assumptions can break down during the later stages of atmospheric growth.

@@

To examine the regimes where our cooling model is valid, we compare the model luminosity to the additional neglected luminosity, $L_{\rm negl}$, that a more detailed model would have generated in the radiative zone.  To compute $L_{\rm negl}$ we compute the entropy difference between successive radiative zone solutions.  We then integrate the energy equation, $\p L / \p m = - T \p S/ \p t$, over the average depth of the radiative zone.\footnote{While useful as a diagnostic, the neglected luminosity cannot reliably correct the global cooling model because where $L_{\rm negl}$ is significant, the structure of the radiative zone would be inaccurate.}

 \Fig{fig:coolingterms} shows that the neglected luminosity is indeed negligible during the early stages of evolution.  However, $L_{\rm negl}$ exceeds the model luminosity, $L$, at high masses, $M_{\rm atm} > 3 M_\oplus$ in this case, meaning that our model is becoming inaccurate.  Nevertheless, our cooling model gives both conservative and reasonably reliable estimates of core accretion timescales.  Our results are conservative -- in the sense of overestimating growth times -- because the neglected luminosity would give more rapid evolution.  Moreover, our results are reasonably accurate because the bulk of growth time is spent at low masses, when $L_{\rm negl}$ is small and safely neglected.

The (non-vanishing) individual terms in the global cooling model of \Eq{eq:coolingglobal}, evaluated at the RCB, are also plotted in \Fig{fig:coolingterms}.  At low masses, the change in energy, $- \dot{E}$, makes the dominant contribution to luminosity.  As the mass increases the surface terms become more significant, led by the accretion energy.  However the surface terms are everywhere smaller than $L_{\rm negl}$.  Thus while the surface terms appear to reduce the total luminosity (compared to $-\dot{E}$), $L_{\rm negl}$ can more than compensate, showing again that our model conservatively underestimates cooling.

At yet higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.  Mathematically, the breakdown manifests as negative timesteps when advancing to a higher atmospheric mass.  This unphysical result has no deep significance, as we already know that the neglected luminosity is large enough to give incorrect results.  Sometimes this breakdown occurs even before the crossover mass is reached.  The main practical effect of this behavior is that we halt the evolution where the model breaks down whenever we cannot reach the crossover mass.   Fortunately, the elapsed time to either the crossover mass or the breakdown is well-behaved and, most importantly, dominated by lower atmospheric masses where the model is accurate.

\begin{figure}[tb]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cooling_a10_Mc5_rcb_paper.pdf}
%\vspace{-0.5in}
\caption{The relative contribution of the various terms in the cooling \Eq{eq:coolingglobal} to the atmosphere luminosity, for a planet forming around a $5 M_\oplus$ core at 10 AU. The $L_{\rm negl}$ curve gives the extra luminosity that would have been generated in the radiative zone, but is not included in our two-layer cooling model. The neglected luminosity is lower than the model luminosity $L$ for low atmosphere mass, and becomes comparable at higher masses ($>3 M_{\oplus}$ in this particular case). As atmosphere growth is slowest at lower masses, neglecting the additional luminosity generated in the radiative layer yields accurate results, to order of magnitude.}
\label{fig:coolingterms}
\end{figure}

>>>>

The lowest mass atmosphere -- which we take as our initial state -- is fully convective with the same entropy as the disk.  In \Fig{fig:profiles} this state is the 5.01 $M_{\oplus}$ solution with no radiative zone.  Atmospheric cooling and contraction allows the accretion of more gas.  In the convective zones, higher mass solutions contain lower entropy and higher pressures.  A radiative zone emerges to connect the lower entropy interior to the higher entropy disk.  \Fig{fig:profiles} shows that this radiative zone is already fairly deep in the 5.10 $M_\oplus$ solution.

The detailed structure of these profiles is well approximated by our non-self-gravitating, analytic solutions.  Deep in convective zones, \Eq{eq:deep} gives that $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$.  This behavior is seen in the low mass solutions in \Fig{fig:profiles}.  As explained by \Eq{eq:TPsg}, the high mass solutions show a slightly flatter profile in $T$ and also in $P \propto T^{1/\delad}$.  In agreement with \Eq{eq:cb2}, the radiative zones remain nearly isothermal, even for the higher masses.  Consequently the pressure increases nearly exponentially with inverse depth.

@@

The cooling model of \S\ref{sec:twolayer} is used to connect solutions of different atmospheric masses into an evolutionary sequence.  \Fig{fig:Ltplot} shows the luminosity evolution and the elapsed time as a function of atmospheric mass for a $5 M_{\oplus}$ core at $60$ AU, i.e.\ the same parameters as in \Fig{fig:profiles}.  During the early stages of atmospheric growth, the luminosity drops sharply.  This behavior is seen in both the full numerical solutions (with solid lines) and the analytic model (with dashed lines).  The increasing pressure depth of the RCB with mass causes increases in the optical depth ($\propto \kappa P)$ and decreases in the radiative luminosity.  This behavior is described in \Eqs{eq:crit}{eq:Lcb}.

At later stages of evolution, the numerical model in \Fig{fig:Ltplot} shows a flat luminosity with increasing mass and also time (not shown).  By contrast the non-self-gravitating analytic model gives a luminosity that continues to drop with increasing mass.  To understand this difference note that the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, holds in both cases.  Thus accounting for the higher enclosed mass in the self-gravitating model gives a somewhat higher luminosity, as desired.  However, the main effect is that \Eq{eq:crit} -- which describes a nearly linear relation between atmospheric mass and RCB pressure --  breaks down for self-gravitating solutions.  This behavior can be seen in the top panel of \Fig{fig:profiles} where the $P\cb$ increases significantly from 5.10 to 6.0 $M_\oplus$, but only increases relatively modestly with further growth to 9.24 $M_\oplus$.   To allow these relatively lower $P\cb$ values (and thus higher luminosities) in the higher mass solutions, the radius $R\cb$ must shift outwards, as is shown.

The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot} is also a direct result of the enhanced cooling (higher luminsoties) in the self-gravitating models.  Clearly self-gravity plays a very important role in atmospheric evolution, even well before the crossover mass.

%Before discussing time evolution (both the elapsed time from the bottom panel of \Fig{fig:Ltplot} and the characteristic growth time of the atmosphere), we briefly discuss the effect of decreasing the atmospheric dust abundance on the atmospheric evolution.  
%AY abo e sentence is bad, we want to discuss science and not "direct traffic"

In \Fig{fig:LtvsMopacity}, the opacity normalization in \Eq{eq:opacitylaw} is reduced by factors of 10 and 100.  Lower opacities result in higher luminosities and faster evolution.  Our model thus confirms this well-established result \citep{HubBod05}.  While clearly an important effect, atmospheric dust opacities are difficult to robustly predict. Ablation of infalling solids is a dust source.  Sinks include the sequestration of solids in the core and dust settling through the radiative zone.  Grain growth both reduces dust opacities per unit mass and favors settling.  Our scenario of negligible ongoing particle accretion tends to favor low dust opacities.   To be conservative, however, our reference case considers full Solar abundances. The effect of opacity reduction on the critical core mass is described in \S\ref{sec:critical}.

@@

\Fig{fig:growthtime} plots the evolution of the atmospheric growth timescale, $M_{\rm atm}/\dot{M}$,  around a $5 M_{\oplus}$ core at several locations in our reference disk model.  This instantaneous growth time shows clearly that the atmosphere spends the bulk of its time growing though intermediate atmospheric masses, $\sim 1 -3$ $M_\oplus$ in this case.  Growth times are short both early -- when the radiative zone is relatively transparent -- and late -- when self-gravity accelerates growth.  The accelerated growth at higher masses is also evident in \Figs{fig:Ltplot}{fig:LtvsMopacity} (bottom panels) as the flattening of the $t$ vs.\ $M_{\rm atm}$ curve.

The fact that growth times peak before the crossover mass, when $M_{\rm{atm}} < M_{\rm c}$, is helpful for understanding atmospheric growth.  The estimated timescale to runaway growth can be well approximated by only considering the time it takes to grow to the crossover mass, or even to a somewhat lower mass.  This insensitivity to the upper mass threshold is characteristic of accelerating growth.  We next show how this behavior is fortuitous, as our model assumptions can break down during the later stages of atmospheric growth.

@@

To examine the regimes where our cooling model is valid, we compare the model luminosity to the additional neglected luminosity, $L_{\rm negl}$, that a more detailed model would have generated in the radiative zone.  To compute $L_{\rm negl}$ we compute the entropy difference between successive radiative zone solutions.  We then integrate the energy equation, $\p L / \p m = - T \p S/ \p t$, over the average depth of the radiative zone.\footnote{While useful as a diagnostic, the neglected luminosity cannot reliably correct the global cooling model because the effects of $L_{\rm negl}$ on the structure of the radiative zone are still ignored.}

 \Fig{fig:coolingterms} shows that the neglected luminosity is indeed negligible during the early stages of evolution.  However, $L_{\rm negl}$ exceeds the model luminosity, $L$, at high masses, $M_{\rm atm} > 3 M_\oplus$ in this case, meaning that our model becomes inaccurate at late states.  Nevertheless, our cooling model gives  reasonably reliable estimates of core accretion timescales.   As shown above, the growth time is mostly spent at low masses, when $L_{\rm negl}$ is small and safely neglected.
 %Our results are conservative -- in the sense of overestimating growth times -- because the neglected luminosity would give more rapid evolution.  
 %AY: not sure we can back up the conservative claim because of the detailed effects on structure that haven't been calculated.

The (non-vanishing) individual terms in the global cooling model of \Eq{eq:coolingglobal}, evaluated at the RCB, are also plotted in \Fig{fig:coolingterms}.  At low masses, the change in energy, $- \dot{E}$, makes the dominant contribution to luminosity.  As the mass increases, the surface terms become more significant, led by the accretion energy.  However the surface terms are everywhere smaller than $L_{\rm negl}$.  Thus wherever our model is accurate -- including the crucial early phases of growth -- surface terms are a minor correction.  
%while the surface terms appear to reduce the total luminosity (compared to $-\dot{E}$), $L_{\rm negl}$ can more than compensate, showing again that our model conservatively underestimates cooling.
%AY:again don't think conservative claim can be so easily justified.

At yet higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.  Mathematically, the breakdown manifests as negative time steps when advancing to a higher atmospheric mass.  This unphysical result has no deep significance, as the neglected luminosity is already large enough to give inaccurate results.  Sometimes this breakdown occurs even before the crossover mass is reached.  The main practical effect of this behavior is that we halt the evolution where the model breaks down whenever we cannot reach the crossover mass.   Fortunately, the elapsed time to either the crossover mass or the breakdown is well-behaved and, most importantly, dominated by lower atmospheric masses where the model is accurate.

\begin{figure}[tb]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cooling_a10_Mc5_rcb_paper.pdf}
%\vspace{-0.5in}
\caption{Individual terms in the atmospheric cooling model of \Eq{eq:coolingglobal}, for a $5 M_\oplus$ core at 10 AU.  The dashed curve for accretion energy indicates a negative contribution.  All quantities are evaluated at the RCB, except for $L_{\rm negl}$, the extra luminosity that would have been generated in the radiative zone, but is neglected in our model. The neglected luminosity is a small correction to the model luminosity $L$ for $M_{\rm atm} \lesssim 3 M_{\oplus}$.   Since these low masses dominate growth times, our model is roughly accurate.}
\label{fig:coolingterms}
\end{figure}
